---
title: "Californian fires HIA"
author: "Ryan Gan"
date: "12/28/2017"
output: html_document
---

## Introduction

HIA formula
delta_y = daily_baseline_rate x (1-e^-beta x delta_smk) x county pop at risk

Things I have:
- beta and population-weighted delta smk
- county pop at risk (easy enough to get and i'll need it for the rate)

## Things I need:
- Daily ED rate for outcomes of interest; i can calculate this from the data I have from colorado, washington, and oregon. 
- Estimate only association between smoke and ED for big time-serieis analysis


In the mean time I'm going to use estimates from this CDC table from 2014.
https://www.cdc.gov/nchs/data/nhamcs/web_tables/2014_ed_web_tables.pdf

All ED visits are estimated at 45.1 per 100 persons per year. Std error is estimated at 3.7 for this rate.

Respiratory ED visits are estimate to make up 10.4% of the ED visits. Std error is estimated at 0.4.

I calculated the daily respiratory ed visit rate to be 1.285 x 10^-4, or 1.285/10000 persons. 

This is not ideal as I have no idea about specific conditions like asthma; I will estimate these from data I have, but for now it will let me add it to the demo.

## Read data

Population-weighted smoke data from 2017-10-11.

```{r setup}
library(tidyverse) 
library(stringr)
library(rgdal) # for testing to write polygon
library(sf)
library(sp)

# knitr options
knitr::opts_chunk$set(fig.width=8, fig.height=6, warning=FALSE, message=FALSE)
```

```{r read_data}
# county path
poly_path <- "./data/example_data/california_county"
poly_layer <- "california_county"
# read county polygons
cali_county <- st_read(dsn = poly_path, layer = poly_layer)
# read bluesky forecast polygon
cali_shape <- readOGR(dsn = poly_path, layer = poly_layer)

# read population weighted smk 
county_smk_pm <- read_csv("./data/example_data/2017-10-11_popwt_pm_county.csv")
# read in california county population
county_pop <- read_csv(paste0("./data/example_data/",
  "2015-california_population.csv")) 
# daily respiratory rate; no asthma rate
resp_rate <- 1.285/10000
# beta for respiratory CR function
resp_beta <- log(1.052)
# read cali polygon

```

Estimation of county respiratory ED visits.

```{r county_resp}
# set up hia dataframe for each county
hia_est <- county_pop %>% 
  mutate(pop_2015 = as.numeric(respop72015)) %>% 
  dplyr::select(FIPS, pop_2015) %>% 
  left_join(county_smk_pm, by = "FIPS") %>% 
  mutate(base_rate = resp_rate,
         beta = resp_beta,
         resp_y = round((base_rate * (1-exp(-(beta)*same_day)) * pop_2015),0))
```

Smoke map.

```{r smoke_map}
# join smoke estimates and HIA estimates to sf file
# this will have to be to a spatial polygon 
cali_county <- cali_county %>% 
  left_join(hia_est, by = "FIPS")

# plot population-weighted smoke
ggplot(cali_county) +
    geom_sf(aes(fill=same_day)) +
    scale_fill_gradient2(name="Pop Wt Smoke PM2.5", 
      low="#03001e", mid="#ec38bc", high="#fdeff9", midpoint = 200) +
  geom_sf(data=cali_county, color = "white", alpha = 0) +
  theme_minimal()
```

HIA map.

```{r hia_map}
ggplot(cali_county) +
    geom_sf(aes(fill=resp_y)) +
    scale_fill_gradient(name="Resp ED", 
      low="#191654", high="#43c6ac") +
  geom_sf(data=cali_county, color = "white", alpha = 0) +
  theme_minimal()

# commenting out and will work this in to daily script
# # join smoke to shapefile; sp merge should work
# cali_poly <- sp::merge(cali_shape, hia_est, by = "FIPS")
# # write spatialy poly
# writeOGR(obj = cali_poly, dsn = "./data/example_data/hia_poly", 
#          layer = "hia_poly", driver = "ESRI Shapefile", overwrite_layer = T)
```

Thoughts:
- Monte-carlo for bounds?
- Future association study should be place of service specific (don't mix Ed and urgent care).

