---
title: "Population-Weighting Tutorial"
author: "Ryan Gan"
date: "12/21/2017"
output: html_document
---

Calculate population-weighted smoke PM2.5 estimates.
Background here.
Smoke from the big fire in Norther California on 2017-10-11.

Enter formula here.
```{r setup}
library(tidyverse) 
library(sf)

# knitr options
knitr::opts_chunk$set(fig.width=8, fig.height=6, warning=FALSE, message=FALSE)
```

Read in the following files: California county sf object, Bluesky grid sf object subsetted to California, Bluesky smoke PM~2.5~ estimates, proportion intersect between counties and bluesky grid cells, and population densities.

```{r read_data}
# read smoke pm estimate for grid 
smk_pm <- read_csv('./data/example_data/2017-10-11_california_bluesky_pm.csv')

# read population estimates for grid
# this if finally right; save it
cal_population <- read_csv(paste0("./data/example_data/",
  "2015-california_grid_population.csv"))

# 4480 rows and 1 column (pm values)
# read sf files 
grid_path <- "./data/example_data/california_bluesky"
grid_layer <- "california_bluesky"
# county path
poly_path <- "./data/example_data/california_county"
poly_layer <- "california_county"

# read in bluesky grid
bluesky_grid <- st_read(dsn = grid_path, layer = grid_layer) %>% 
  dplyr::select(-layer_1) %>% 
  dplyr::left_join(smk_pm, by = c("id"="grid_id")) %>% 
  dplyr::left_join(cal_population, by = "id")
  # cap at 250, taking out for now 
  # dplyr::mutate(pm = ifelse(pm > 250, 250, pm))
# read county polygons
cali_county <- st_read(dsn = poly_path, layer = poly_layer)

# read proportion intersect
grid_county_pi <- read_csv("./data/example_data/cali_bluesky_pi.csv",
    col_types = cols(.default = col_character())) %>% 
  mutate_all(as.numeric)
```

Plot of smoke during October 11th. There are some cells with very high values that I usually cap for display purposes.

```{r smoke_plot}
# visualize
ggplot(bluesky_grid) +
  geom_sf(aes(fill=pm)) +
    scale_fill_gradient2(name="Smoke PM2.5", 
      low="#03001e", mid="#ec38bc", high="#fdeff9", midpoint = 400) +
  geom_sf(data=cali_county, color = "white", alpha = 0) +
  theme_minimal()
# looks okay
```

Let's see where most people live in the state of California.

```{r popdensity_map}
# plot popluation density to see if it looks correct
ggplot(bluesky_grid) +
  geom_sf(aes(fill = popden)) +
    scale_fill_gradient2(name="Population Density per km^2", 
      low="#2c3e50", mid="#00c9ff", high="#92fe9d", midpoint = 2000) +
  geom_sf(data=cali_county, color = "white", alpha = 0) +
  theme_minimal()
# that looks much better
```

Now we'll calculate population-weighted PM~2.5~, which as the name suggests, weights smoke values in each county based on where most people live.

Setting up matrices and vectors for some linear algebra.
```{r population_wt}
# proportion intersect between county and grid matrix
pi_mat <- as.matrix(grid_county_pi[,2:59])
rownames(pi_mat) <- grid_county_pi$grid_id



# pm value vector (duplicating to two rows to test the next day)
pm_mat <- as.matrix(cbind(smk_pm$pm, smk_pm$pm))
# population density value vector
popden <- cal_population$popden
# multiply population vector by pm vector
pm_pop_mat <- popden * pm_mat

# multiply pm population matrix by zip population vector to get a summed
# estimate of pm exposure that day
county_grid_pm_mat <- t(pi_mat) %*% pm_pop_mat

# estimate the population density per county
popden_county <- t(pi_mat) %*% popden

# calculate the inverse of the popden_couty matrix
popden_county_inverse <- 1/popden_county

# population weighted estimates for each county for same-day and next-day
# multiply by a vector of the inverse population density for each county
# final product
county_pop_wt_smk <- county_grid_pm_mat * as.vector(popden_county_inverse) 

# create dataframe
pm_county_wt <- as_data_frame(county_pop_wt_smk) %>% 
  rename(same_day = V1, next_day = V2) %>% 
  mutate(FIPS = as.character(stringr::str_sub(rownames(county_pop_wt_smk), 
                                              start = 6L)))
```

County population-weighted smoke PM~2.5~ estimates.

```{r popwt_plot}
# join pop wt county smoke estimates
cali_county <- cali_county %>% 
  left_join(pm_county_wt, by = "FIPS")

# plot population-weighted smoke
ggplot(cali_county) +
    geom_sf(aes(fill=same_day)) +
    scale_fill_gradient2(name="Pop Wt Smoke PM2.5", 
      low="#03001e", mid="#ec38bc", high="#fdeff9", midpoint = 200) +
  geom_sf(data=cali_county, color = "white", alpha = 0) +
  theme_minimal()
```

Now I need to work on the HIA component of this.