---
title: "Calculating Proportion Intersect using Simple Features (sf)"
author: "Ryan Gan"
date: "10/12/2017"
output: html_document
---

## Introduction 

This document focuses on calculation of the intersection between two spatial polygons (shapefiles). The reason I want to calculation the intersection is so that I could population weight PM~2.5~ values based on grid cells for a given politcal boundary, in this case, county. This is my first time taking the "sf" package for a go and I'm excited about it's potential. 

## Reading in county shapefile and grid as simple features (sf) 

This is example code on using the relatively new "sf" simple features package. Perviously, I have used the sp, rgeos, and rgdal package for many spatial tasks/calculations. Use of sf allows for a faster and more memory-efficient way to perform spatial tasks. From what I can tell, the major difference is that a simple feature is a dataframe with information on the geometry as part of the observation, rather than a spatial polygon object. This would explain why it's more memory efficient. Another very attractive feature is that "sf" follows tidy principles for data manipulation and managment, and is compatible with the tidyverse. For this, I use the development version of ggplot2 to access the geom_sf feature. I expect "sf" and tidyverse to work even better together in the future.

Let's read in the shapefile polygons as simple features for the Bluesky smoke forecasting grid and the polygons for United States counties (excluding Alaska and Hawaii).

```{r setup}
# trying out sf for the first time
library(sf)
# note that you need the development version of ggplot2 
library(tidyverse)
# load wrfgrid polygon ----
# define relative path to polygon file
poly_path <- "./data/bluesky_grid"
poly_layer <- "bluesky_grid"
# county path
co_path <- "./data/us_county"
co_layer <- "us_county"

# read in bluesky grid
bluesky_grid <- st_read(dsn = poly_path, layer = poly_layer)
# the bluesky grid does not have an ID so we will assign each cell a number
bluesky_grid$id <- as.numeric(seq(1:94068))
# read county polygons
us_county <- st_read(dsn = co_path, layer = co_layer)
```

Now that we've read in the two polygons, it looks like there are some unique features to the sf package vs some other packages like rgdal or rgeos. It looks like the imported simple features object contains more information than the simple polygon, which is cool. I haven't plotted the entire US county shapefile and the Bluesky grid because it's hard to see, but with smaller shapefiles and grids, I usually do this as a check.

## Subsetting specific polygons (California)

In the past, we calculated the proportion of intersection between each county polygon and grid cell for one state, which worked fine. However, when we try to apply the same method to the entire US, we ran in to what I think were memory problems (stack overflow?), and we were never able to get our desired product, a matrix of the proportion intersect for each cell id and county id. This was on a relatively decent high-performance computer too running in parallel and definately wouldn't work on your standard desktop or laptop. To solve this problem, we'll subset each state and only the cell grids of that state, which should reduce our memory use substantially. We can save the matrix for each state, free up the memory used, and start again with a new state. We can then combine all the matrices from each state to get our desired matrix. 

For this example, we'll take it to a much smaller scale: the state of California.

We'll also check the plots of each object to make sure we have what we need.

```{r california_subset}
# subset to california by FIPS code "06"
cali_county <- us_county[us_county$STATEFP=="06", ] %>% 
  # making a FIPS ID that I'll use later
  mutate(FIPS = paste0(STATEFP, COUNTYFP))

# clip by bbox function ------
bbox_clip <- function(sf, bbox) {
  # find the CRS of the sf object
  crs <- sf::st_crs(sf)$proj4string
  # create matrix
  x <- c(bbox[1], bbox[1], bbox[3], bbox[3], bbox[1])
  y <- c(bbox[2], bbox[4], bbox[4], bbox[2], bbox[2])
  coords <- matrix(cbind(x, y), ncol=2)
  # create polygon and assign same coord crs as sf object
  coords_poly <- sp::Polygon(coords)
  bbox_poly <- sp:: SpatialPolygons(list(sp::Polygons(list(coords_poly),
    ID = "bbox")), proj4string = sp::CRS(crs))
  # convert to sf feature
  bbox_sf <- st_as_sf(bbox_poly)
  # clip sf object
  clipped_sf <- sf[bbox_sf,]
  return(clipped_sf)
}

cali_bbox <- st_bbox(cali_county)

# subset a smaller bluesky grid
cali_grid <- bbox_clip(bluesky_grid, cali_bbox)

# plot
ggplot(cali_county) + 
  geom_sf() +
  geom_sf(data=cali_grid, aes(), alpha = 0.0) +
  theme_bw()

# write the two sf files for california to save for example
# st_write(cali_county, paste0("./data/example_data/",
#   "california_county/california_county.shp"))
# 
# st_write(cali_grid, paste0("./data/example_data/california_bluesky/",
#                            "california_bluesky.shp"))
```

Looks like we have all the grids we need for the state of California. Let's move on to the proportion of intersection between each county and grid in each state. 

## Proportion intersect calcuations for Los Angeles county

Before we apply a function to the whole state, we'll subset start small with one county and grids in that county. I do this for a couple reasons. The first is I can test if my code is working on a small subset on my laptop before moving it to the server. The second is that it is easy to scale it up and use *apply* functions to apply the formula for each grid and county (for the entire US, eventually). The last reason is that I'll check the final matrix for the grid ids for that particular county to make sure they line up with these calculations here.

I'm going to pick Los Angeles county since it's small and has some islands. We can subset in the same way we did before.

```{r la_subset, warning = F}
# subset orange county by fips code
la_county <- cali_county[cali_county$COUNTYFP=="037",]
# subset grids to orange county
la_grid <- cali_grid[la_county, ]

# dataframe of name and centroid to plot
grid_names <- la_grid %>% 
  group_by(id) %>% 
  mutate(lon = unlist(st_centroid(geometry))[1],
         lat = unlist(st_centroid(geometry))[2]) %>% 
  ungroup() %>% 
  select(id, lon, lat)
# convert simple featrues to dataframe
st_geometry(grid_names) <- NULL

# plot grid over LA county and label grid cells
ggplot() + 
  geom_sf(data = la_county) +
  geom_sf(data=la_grid, alpha = 0.0) +
  geom_text(data=grid_names, aes(x=lon, y=lat, label=id), size = 2.5,
            angle=45)+
  theme_bw()
```

We'll calculate the proportion intersection of one grid, 55299, for the upper part of LA county.

```{r proportion_intersect, warning=F}
# first find the intersection between the two shapes
intersect_sf <- st_intersection(la_county, la_grid[1,])
# find the proportion intersect by subtracting the area of the intersection
# by the area of the grid cell. as.numeric ditches the unit
prop_intersect <- as.numeric(st_area(intersect_sf)/st_area(la_grid[1,]))
# proportion of LA county in the grid 55299, which is the upper left hand corner 
prop_intersect
```

Since "sf" brings in polygons as simple features as a data frame instead of multiple little polygons as part of a larger polygon, that changes how I can calculate proportion intersect. I believe it's more intuitive, takes less memory, and much faster. 

Calculating the proportion intersect between all grids in LA county with LA county is as simple as putting the whole grid in the intersect function. Before, I had to subset each grid and find each intersection with the county polygon, which took a lot longer.

```{r la_intersection, warning=F}
# sf might make this easier since each object comes in as a dataframe and not
# a bunch of little polygons as part of a big polygon
# find intersection of grid for the county
grid_area <- st_intersection(st_geometry(la_grid),st_geometry(la_county))
# caluclate proportion intersect 
prop_int <- as.numeric(st_area(grid_area)/st_area(la_grid))
# now I need to assign it back to the county shapefile
la_grid$proportion <- prop_int

# make a dataframe with rounded proportion values
grid_names <- la_grid %>% 
  mutate(proportion = round(prop_int,2)) %>% 
  group_by(id) %>% 
  mutate(lon = unlist(st_centroid(geometry))[1],
         lat = unlist(st_centroid(geometry))[2]) %>% 
  ungroup() %>% 
  select(id, lon, lat, proportion)
# convert simple featrues to dataframe
st_geometry(grid_names) <- NULL

# plot
ggplot(la_county) + 
  geom_sf() +
  geom_sf(data=la_grid, aes(fill=proportion), alpha=0.7) +
  geom_text(data=grid_names, aes(x=lon, y=lat, label=proportion), 
            size = 2.5, angle=45) +
  theme_bw()
```

I calculated the intersection for each grid with the county shape, plotted a gradiet fill and the value of the proportion intersect. Based on the eye-ball test, it looks like our function worked. Our top-left grid cell also matches what we calculated before (I rounded to the nearest hundreth).

Try it out on the state of California. I don't think I can use purrr map functions yet. I'll use a for loop.

```{r california_intersect, warning=F}
# I had some problems feeding the sf object directly in to the function when I
# tried to use the apply or map functions, which led me to think purrr and map
# were not compatible yet. If I feed in a vector of the state/county fips,
# this should work

# tibble of just the grid_id to join data too
prop_int_tibble <- cali_grid$id %>% 
  tibble() %>% 
  rename(grid_id = ".")

start_time <- Sys.time()

for(i in 1:length(cali_county$FIPS)){
  # subset county to find intersect
  county <- slice(cali_county, i)
  # extract fips number for variable name
  fips_id <- paste0("fips_", county$FIPS)
  # subset grid cells that touch any part of the county
  grid <- cali_grid[county,]
  # subset the intersected polygon
  inter_poly <- st_intersection(grid, county) %>% 
    # filter only to polygon or multipolygon type 
    # to avoid errors with point or line types
    filter(st_is(., c("POLYGON", "MULTIPOLYGON"))==T)
  # filter grid ids to only grids in the inter_poly object
  grid2 <- grid %>% filter(id %in% inter_poly$id) 
  # find proportion intersect with original grid
  prop_int <- as.numeric(st_area(inter_poly)/st_area(grid2))
  # subset grid id
  grid_id <- grid2$id
  # make a tibble
  county_grid_int <- tibble(grid_id, prop_int) %>% 
    set_names(c("grid_id", fips_id))
  # join with full tibble
  prop_int_tibble <- prop_int_tibble %>% 
    left_join(county_grid_int, by = "grid_id")
} # end loop

# stop cluster
#stopCluster(cl)

stop_time <- Sys.time()
compute_time <- stop_time - start_time
# run time
compute_time

# set missing NA values to 0
cali_bluesky_pi <- prop_int_tibble %>% 
  mutate_all(funs(replace(., is.na(.), 0)))

# save file for a vignette
write_csv(cali_bluesky_pi, paste0("./data/example_data/cali_bluesky_pi.csv"))
```

Checking LA county to see if estimates are similar.

```{r la_check}
la_county_tibble <- prop_int_tibble %>% 
  select(grid_id,fips_06037) %>% 
  # remove missing values
  filter(!is.na(fips_06037))

all.equal(la_county_tibble, la_grid)

summary(la_county_tibble)
summary(la_grid)
```

Next steps would be to figure out how to use the sf package with apply or map type functions. I think the for loop will work and so far apply and map don't work and I need to move on. I've saved the code I do have for now.

```{r purrr attempt, warning = F, eval = F}
proportion_intersect <- function(poly_sf, poly_id, grid_sf, grid_id){
  # enquo lazy eval
  poly_id <- enquo(poly_id)
  grid_id <- enquo(grid_id)
  # subset grid sf that contains poly_i
  grid_i <- grid_sf[poly_sf,]
  # set grid_i crs to poly_sf crs
  standard_crs <- st_crs(poly_sf)
  st_crs(grid_i) <- standard_crs 
  # assume grid_i and poly_sf are constant
  st_agr(poly_sf) = "constant"
  st_agr(grid_i) = "constant"
  # proportion intersect
  intersect_sf <- st_intersection(grid_i, poly_sf) %>% 
    # filter only to polygon or multipolygon type 
    # to avoid errors with point or line types
    filter(st_is(., c("POLYGON", "MULTIPOLYGON"))==T)
  # calculation of proportion intersect
  proportion <- as.numeric(st_area(intersect_sf)/st_area(grid_i)) %>% 
    data_frame() %>% rename(proportion = ".")
  # column bind the proportion to the intersect sf object
  output_df <- intersect_sf %>% 
    # eventually replace these with generic names
    dplyr::select(!!grid_id, !!poly_id) %>% 
    bind_cols(proportion)
  # remove geometry
  st_geometry(output_df) <- NULL
  return(output_df)
}

?st_crs
test <- cali_county %>% nest


check <- proportion_intersect(poly_sf = test, poly_id = FIPS, 
                              grid = bluesky_grid, grid_id = id)

# purrr attempt
prop_int_purrr <- test %>% 
  lmap(~proportion_intersect(poly_sf = ., poly_id = FIPS, 
                            grid = bluesky_grid, grid_id = id))

apply(test,1, function(x) st_area(x))


map(test, function(x) st_area(x))

st_crs(test)
st_crs(bluesky_grid)

# extracting population densities from bonne's bluesky grid -----
# i will move it later

pop_nc <- ncdf4::nc_open("./data/blueskypopulation.nc")

cali_id <- bluesky_grid$id
# extract population and population density for california grid cells
pop <- as.vector(ncdf4::ncvar_get(pop_nc, varid = "Population"))
popden <- as.vector(ncdf4::ncvar_get(pop_nc, varid ="PopulationDensity"))

# extract latlon
lat <- ncdf4::ncvar_get(pop_nc, varid ="latitude")
lon <- ncdf4::ncvar_get(pop_nc, varid = "longitude")
# expand grid
lonlat <- as.matrix(expand.grid(lon,lat))

# create population dataframe and add names
population_df <- data.frame(cbind(lonlat, pop, popden))
# assign names
names(population_df) <- c("lon", "lat", "pop", "popden")

# sf label starts top left and goes right, then down one row
# sort by desc(lat) then lon to match how i labeled the sf objects
population_df <- population_df %>%  
  arrange(desc(lat), lon) %>% 
  mutate(id = seq(1:94068))

# now I think i can subset to california by row number
cal_population <- population_df %>% 
  filter(id %in% cali_id) %>% 
  dplyr::select(id, lon, lat, pop, popden)

# this if finally right; save it
write_csv(cal_population, paste0("./data/example_data/",
                                 "2015-california_grid_population.csv"))

```



