---
title: "Calculating Proportion Intersect using Simple Features (sf)"
author: "Ryan Gan"
date: "10/12/2017"
output: html_document
---

## Introduction and setup

This is example code on using the relatively new "sf" simple features package. Perviously, I have used the sp, rgeos, and rgdal package for many spatial tasks/calculations. Use of sf allows for a faster and more memory-efficient way to perform spatial tasks. From what I can tell, the major difference is that a simple feature is more like a dataframe rather than a spatial polygon object. This would explain why it's more memory efficient. Another very attractive feature is that "sf" follows tidy principles for data manipulation and managment, and is compatible with the tidyverse. For this, I use the development version of ggplot2 to access the geom_sf feature. I expect "sf" and tidyverse to work even better together in the future.

This document focuses on calculation of the intersection between two spatial polygons (shapefiles). The reason I want to calculation the intersection is so that I could population weight PM~2.5~ values based on grid cells for a given politcal boundary, in this case, county. This is my first time taking the "sf" package for a go and I'm excited about it's potential. 

Let's read in the shapefile polygons for the Bluesky smoke forecasting grid and the polygons for United States counties (excluding Alaska and Hawaii).

```{r setup}
# trying out sf for the first time
library(sf)
# note that you need the development version of ggplot2 
library(tidyverse)
# load wrfgrid polygon ----
# define relative path to polygon file
poly_path <- "../data/bluesky_grid"
poly_layer <- "bluesky_grid"
# county path
co_path <- "../data/us_county"
co_layer <- "us_county"

# read in bluesky grid
bluesky_grid <- st_read(dsn = poly_path, layer = poly_layer)
# the bluesky grid does not have an ID so we will assign each cell a number
bluesky_grid$id <- as.numeric(seq(1:94068))
# read county polygons
us_county <- st_read(dsn = co_path, layer = co_layer)
```

Now that we've read in the two polygons, it looks like there are some unique features to the sf package vs some other packages like rgdal or rgeos. It looks like the imported simple features object contains more information than the simple polygon, which is cool. I haven't plotted the entire US and the Bluesky grid because it's hard to see, but I usually do this as a check.

## Subsetting specific polygons

In the past, we calculated the proportion of intersection between each county polygon and grid cell for one state, which worked fine. However, when we try to apply the same method to the entire US, we ran in to what I think were memory problems (stack overflow?), and we were never able to get our desired product, a matrix of the proportion intersect for each cell id and county id. This was on a relatively decent high-performance computer too running in parallel and definately wouldn't work on your standard desktop or laptop. To solve this problem, we'll subset each state and only the cell grids of that state, which should reduce our memory use substantially. We can save the matrix for each state, free up the memory used, and start again with a new state. We can then combine all the matrices from each state to get our desired matrix. 

For this example, we'll take it to a much smaller scale: the state of California.

We'll also check the plots of each object to make sure we have what we need.

```{r california subset}
# subset to california by FIPS code "06"
cali_county <- us_county[us_county$STATEFP=="06", ] %>% 
  # making a FIPS ID that I'll use later
  mutate(FIPS = paste0(STATEFP, COUNTYFP))
# subset the bluesky grid to  grid cells that intersect a california county
cali_grid <- bluesky_grid[cali_county, ]

# plot
ggplot(cali_county) + 
  geom_sf() +
  geom_sf(data=cali_grid, aes(), alpha = 0.0) +
  theme_bw()
```

Looks like we have all the grids we need for the state of California. Let's move on to the proportion of intersection between each county and grid in each state. 

## Proportion intersect calcuations

Before we apply a function to the whole state, we'll subset start small with one county and grids in that county. I do this for a couple reasons. The first is I can test if my code is working on a small subset on my laptop before moving it to the server. The second is that it is easy to scale it up and use *apply* functions to apply the formula for each grid and county (for the entire US, eventually). The last reason is that I'll check the final matrix for the grid ids for that particular county to make sure they line up with these calculations here.

I'm going to pick Los Angeles county since it's small and has some islands. We can subset in the same way we did before.

```{r la county subset, warning = F}
# subset orange county by fips code
la_county <- cali_county[cali_county$COUNTYFP=="037",]
# subset grids to orange county
la_grid <- cali_grid[la_county, ]

# dataframe of name and centroid to plot
grid_names <- la_grid %>% 
  group_by(id) %>% 
  mutate(lon = unlist(st_centroid(geometry))[1],
         lat = unlist(st_centroid(geometry))[2]) %>% 
  ungroup() %>% 
  select(id, lon, lat)
# convert simple featrues to dataframe
st_geometry(grid_names) <- NULL

# plot grid over LA county and label grid cells
ggplot() + 
  geom_sf(data = la_county) +
  geom_sf(data=la_grid, alpha = 0.0) +
  geom_text(data=grid_names, aes(x=lon, y=lat, label=id), size = 2.5, angle=45)+
  theme_bw()
```

We'll calculate the proportion intersection of one grid, 55299, for the upper part of LA county.

```{r proportion intersect, warning=F}
# first find the intersection between the two shapes
intersect_sf <- st_intersection(la_county, la_grid[1,])
# find the proportion intersect by subtracting the area of the intersection
# by the area of the grid cell. as.numeric ditches the unit
prop_intersect <- as.numeric(st_area(intersect_sf)/st_area(la_grid[1,]))
# proportion of LA county in the grid 55299, which is the upper left hand corner 
prop_intersect
```

Since "sf" brings in polygons as simple features as a data frame instead of multiple little polygons as part of a larger polygon, that changes how I can calculate proportion intersect. I believe it's more intuitive, takes less memory, and much faster. 

Calculating the proportion intersect between all grids in LA county with LA county is as simple as putting the whole grid in the intersect function. Before, I had to subset each grid and find each intersection with the county polygon, which took a lot longer.

```{r intersection, warning=F}
# sf might make this easier since each object comes in as a dataframe and not
# a bunch of little polygons as part of a big polygon
# find intersection of grid for the county
grid_area <- st_intersection(st_geometry(la_grid),st_geometry(la_county))
# caluclate proportion intersect 
prop_int <- as.numeric(st_area(grid_area)/st_area(la_grid))
# now I need to assign it back to the county shapefile
la_grid$proportion <- prop_int

# make a dataframe with rounded proportion values
grid_names <- la_grid %>% 
  mutate(proportion = round(prop_int,2)) %>% 
  group_by(id) %>% 
  mutate(lon = unlist(st_centroid(geometry))[1],
         lat = unlist(st_centroid(geometry))[2]) %>% 
  ungroup() %>% 
  select(id, lon, lat, proportion)
# convert simple featrues to dataframe
st_geometry(grid_names) <- NULL

# plot
ggplot(la_county) + 
  geom_sf() +
  geom_sf(data=la_grid, aes(fill=proportion), alpha=0.7) +
  geom_text(data=grid_names, aes(x=lon, y=lat, label=proportion), 
            size = 2.5, angle=45) +
  theme_bw()
```

I calculated the intersection for each grid with the county shape, plotted a gradiet fill and the value of the proportion intersect. Based on the eye-ball test, it looks like our function worked. Our top-left grid cell also matches what we calculated before (I rounded to the nearest hundreth).

Try it out on the state of California. I don't think I can use purrr map functions yet. I'll use a for loop.

```{r grid intersect for cali for loop, warning=F}
# I had some problems feeding the sf object directly in to the function when I
# tried to use the apply or map functions, which led me to think purrr and map
# were not compatible yet. If I feed in a vector of the state/county fips,
# this should work

# tibble of just the grid_id to join data too
prop_int_tibble <- cali_grid$id %>% 
  tibble() %>% 
  rename(grid_id = ".")

# turning this in to a for loop because i'm annoyed
for(i in 1:length(cali_county$FIPS)){
  # subset county to find intersect
  county <- slice(cali_county, i)
  # extract fips number for variable name
  fips_id <- paste0("fips_", county$FIPS)
  # subset grid cells that touch any part of the county
  grid <- cali_grid[county,]
  # output grid IDs
  grid_id <- grid$id
  # subset the intersected polygon
  inter_poly <- st_intersection(st_geometry(grid),st_geometry(county))
  # find proportion intersect with original grid
  prop_int <- as.numeric(st_area(inter_poly)/st_area(grid))
  # make a tibble
  prop_int_tibble <- tibble(grid_id, prop_int) %>% 
    set_names(c("grid_id", fips_id)) %>% 
    right_join(prop_int_tibble, by = "grid_id")
} # end loop

# that was fast enough and easier to follow
summary(prop_int_tibble)

# plot
ggplot(cali_county) + 
  geom_sf() +
  geom_sf(data=cali_grid, aes(), alpha = 0.0) +
  theme_bw()
```

Next steps would be to figure out how to use the sf package with apply or map type functions.

```{r purrr attempt, warning = F, eval = F}
# define intersect function to output a vector 
prop_int_fun <- function(x){
  county <- x
  # extract fips number for variable name
  fips_id <- paste0("fips_", county$FIPS)
  # subset grid cells that touch any part of the county
  grid <- cali_grid[county,]
  # output grid IDs
  grid_id <- grid$id
  # subset the intersected polygon
  inter_poly <- st_intersection(st_geometry(grid),st_geometry(county))
  # find proportion intersect with original grid
  prop_int <- as.numeric(st_area(inter_poly)/st_area(grid))
}

# test of la county
la_test <- prop_int_fun(la_county)

y <- cali_county %>% 
  group_by(FIPS) %>% 
  nest



test1 <- lapply(1:nrow(cali_county), prop_int_fun)


test <- sapply(x$geometry, prop_int_fun)


test3 <- cali_county %>% 
  group_by(FIPS) %>% 
  lmap(prop_int_fun)

test_apply <- lapply(cali_county, prop_int_fun)
```



